Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-29 paper/gd1.tex
words and model

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.\footnote{This transformation is implemented using the
\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
\package{Gala} \texttt{Python} package (\citealt{gala}).}
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream coordinate system, and correct the proper motions for solar
reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
10)~\textrm{kpc}$.
For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
$|\phi_2| < XX^\circ$ in solar-reflex-corrected proper motion components in the
GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
term) and $\mu_{\phi_2, \odot}$.
GD-1 member stars are visible as the over-density at approximately
$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (XX, YY)~\masyr$.

@@

To improve the contrast of the stream over the background, we next cross-match
the sample to the \pans\ photometric catalog and use the photometry to further
clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the distribution of
proper-motion-selected candidate stream stars in a de-reddened \pans\
color-magnitude diagram (CMD).
\todo{Did you also cut on $phi_2$?}.
The over-plotted isochrone is from the MESA Isochrones \& Stellar Tracks (MIST;
\citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.XX$ at a distance of $XX~\kpc$.
We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the \todo{shaded} region
in the bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
stream member candidates after selecting on both proper motions and photometry.
\todo{Comment on how this is the bomb ass diggity and the clearest view of the
stream so far}
\todo{Mention the under-densities}
\todo{Already mention the spur here}

@@

- properties of GD-1 along the track, and beyond

@@

- for study of global properties, we first find stream ridge points as a median $\phi_2$ every $x^\circ$ along $\phi_1$
- then we fit a second order polynomial to these ridge points, and define a stream track as the region $0.6^\circ$ above and beyond the best-fit polynomial (shown with two blue lines in Figure~\ref{fig:track-and-model})

- first measure density variations in background subtracted profile
- find maximum -- morphology as expected from a progenitor in the final stages of dissolution (cit)
- several prominent gaps: density contrast, width
- some (all?) seen in earlier data
- but not related to dust
All of the clearly identifiable portions of the GD-1 stream are located at high
Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
dust extinction or variations in extinction along the stream.
\figurename~\ref{fig:track-and-model} (top panel) shows the $V$-band extinction in the
region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
extinction map (\cite{Schlegel:1998}; hereafter SFD).
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
0.03~\textrm{mag}$.
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable.

Another \todo{segue to this paragraph...}.
However, we find that \todo{...}.
We discuss the scanning pattern in more detail in
Appendix~\ref{sec:completeness}.

\todo{segue sentences...}

From the combined proper motion and CMD selection of stream members
(\sectionname~\ref{sec:data}), it is now clear that the GD-1 stream extends
at least $90^\circ$ in apparent length.
Several clear under-densities and gaps are also
\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with several features

@@

- also show median properties along the stream 5deg window, 1deg spacings
- coordinates and proper motion vary
- these observations are reproduced with a simple model

% \todo{segue sentences...}

% Another \todo{segue to this paragraph...}.
% However, we find that \todo{...}.
% We discuss the scanning pattern in more detail in
% Appendix~\ref{sec:completeness}.

@@

>>>>

\newcommand{\sectionname}{Section}
\newcommand{\equationname}{Equation}
\renewcommand{\tablename}{Table}

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.\footnote{This transformation is implemented using the
\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
\package{Gala} \texttt{Python} package (\citealt{gala}).}
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream coordinate system, and correct the proper motions for solar
reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
10)~\textrm{kpc}$.
For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
$|\phi_2| < XX^\circ$ in solar-reflex-corrected proper motion components in the
GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
term) and $\mu_{\phi_2, \odot}$.
GD-1 member stars are visible as the over-density around
$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog and use the photometry to further clean
the sample.
\figurename~\ref{fig:selection} (bottom right) shows the distribution of
proper-motion-selected candidate stream stars in a de-reddened \pans\
color-magnitude diagram (CMD).
\todo{AB: Did you also cut on $phi_2$?}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.XX$ at a distance of $XX~\kpc$.
We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
stream member candidates after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of individual stars (the
positions are not binned); this is the most pure view of the stream to date.
The increased contrast shows that the stream extends at least another $20^\circ$
to negative longitudes.
Several of the under-densities and gaps hinted at from photometric selection
alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
this significantly cleaner density map of the GD-1 stream.
Finally, at least two new features are visible in this new view of the stream:
(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
-10^\circ$.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.

@@

We extract high-confidence stream stars from the proper motion and CMD selected
sample by further selecting stars close to the main stream track.
We use these main stream track stars to study the global properties of the
stream, and find interesting substructures off of the main stream track.

@@

To study of global properties of the stream as a function of stream longitude,
we extract stream stars around a low-order polynomial model for the median
stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
In detail, we find stream ridge points by computing a running median in $\phi_2$
in steps of $XX^\circ$ along $\phi_1$, then fit a second-order polynomial to the
ridge points to find $\phi_{2, \textrm{track}}(\phi_1)$.
We define the stream as the region within $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.6^\circ$.
\figurename~\ref{fig:XX} (XX panel) again shows the high-confidence
stream stars, and the two curved lines (blue) show the upper and lower boundary
of the stream region.

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{???.pdf}
\end{center}
\caption{%
\todo{???}
\label{fig:XX}
}
\end{figure}

The clearly identifiable portions of the GD-1 stream are located at high
Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
dust extinction or variations in extinction within the stream region.
\figurename~\ref{fig:XX} (XX panel) shows the $V$-band extinction
in the region around the GD-1 stream, computed from the
Schlegel-Finkbeiner-Davis extinction map (\cite{Schlegel:1998}; hereafter SFD).
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$.
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.

We extract properties of the stream from the stream region defined above as a
function of $\phi_1$ by computing properties in overlapping windows:
we set the window size to $XX^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ along the stream region.
\figurename~\ref{fig:track-and-model} shows the median latitude, $\phi_2$,
stream surface density, $XX$, stream width, $XX$, and median proper motions,
$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way.
For the surface density, we subtract an estimate of the background surface
density by \todo{AB}.
To estimate the stream width, $w$, we compute the median absolute deviation,
$\textrm{MAD}$, and use this to estimate a robust standard deviation, $w = 1.5
\times MAD$.

\todo{Comment on density variations, stream width variations. Do we want a
$\Delta \phi_2$ plot (relative to mean stream track) with 15th/85th
percentiles?}

- first measure density variations in background subtracted profile
- find maximum -- morphology as expected from a progenitor in the final stages of dissolution (cit)
- several prominent gaps: density contrast, width
- some (all?) seen in earlier data
- but not related to dust

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
In detail, during simulation, star particles are released from the Lagrange
points of the cluster with a spread in position and velocity set by the mass of
the satellite and the location within the host galaxy, with some tunable scale
parameters.
We use the parametrization and adopted scale parameters of \citet{Fardal:2015},
who fit the release distribution parameters by comparing to full $N$-body
simulations across a range of mass and orbital parameters.
This method can accurately reproduce the mean track of the resulting stream, but
the density of stars along the stream depends strongly on the mass-loss history
and internal kinematics of the progenitor system.
Here we assume a constant mass-loss rate, and release stream particles uniformly
in time from each Lagrange point.

@@

@@

% From the combined proper motion and CMD selection of stream members
% (\sectionname~\ref{sec:data}), it is now clear that the GD-1 stream extends
% at least $90^\circ$ in apparent length.
% Several clear under-densities and gaps are also
% \figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with several features
